---
title: "WG_vignette"
output:
  html_document:
    theme: united
    toc: yes
vignette: |
  %\VignetteIndexEntry{WG_vignette} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Esearch3D  
This vignette introduces how to use Esearch3D for the prediction of the activity of intergenic fragments. 
It goes through all the main operations of the software showing a real application of the tool. Precisely, this vignette replicates the analysis of the mouse embryonic stem cells (mESCs) DNaseI capture Hi-C dataset presented in the publication.  

# Enviroment set up
Let us clean the R environment, set the working directory, load the software package, set the random seed generator for reproducing always the same results and define the number of cores available in the computer to run the operations.
Be careful: set up the number of cores based on your resources. If you are not sure, then just set it to 2

```{r setup}

#Clean workspace and memory ----
rm(list=ls())
gc()

#Set working directory ----
gps0=getwd()
gps0=paste(gps0,"/%s",sep="")
rootDir=gps0
setwd(gsub("%s","",rootDir))

#Load libraries ----
suppressWarnings(suppressMessages(
  library("Esearch3D", quietly = T)
  )
)

#Set variables ----
#Set seed to get always the same results out of this vignette
set.seed(8)
#Set number of cores to parallelize the tasks
n_cores=5
```

# Data set up
For this vignette, we created a DNaseI-capture HiC derived network, whereby captured regions are DNaseI sensible regions from mouse embryonic stem cells (mESC). A  chromatin fragment is represented as a node; a fragment-fragment interaction as an edge. We then integrated genes as nodes within the network. In this case, genes are connected by edges to the fragments situated on the gene.

```{r loading of wg dataset}

#Load and set up the example data ----
data("wg_data_l")
#gene - fragment interaction network generated from DNase_Prop1_mESC_TSS interactions data
gf_net=wg_data_l$gf_net
#gene-fragment-fragment interaction network generated from mESC_DNase_Net interactions data
ff_net=wg_data_l$ff_net
#sample profile with starting values for genes and fragments generated from mESC_bin_matrix_Prop1
input_m=wg_data_l$input_m
#length of chromosomes
chr_len=wg_data_l$chr_len
#gene annotation
ann_net_b=wg_data_l$ann_net_b
#genes of interests
gene_in=wg_data_l$gene_in

```

# Multi-gene two-step propagation
In the first propagation, the gene expression is propagated from their nodes into their directly connected genic fragments. 
In the second propagation, the gene expression is then propagated to the rest of the CIN.

Be careful: `r` is the isolation parameter, which reflects the percentage of information that is kept in the source nodes when propagating. In the first step we want the maximum amount of information to be propagated, so we use a low value. In the second step we can use higher values.
The genic and intergenic fragments receive an imputed activity score (IAS) reflecting the likelihood of enhancer activity.

```{r two step propagation with random walk with restart}

#Two step propagation -----
#Propagated for the network gene-fragment
gf_prop=rwr_OVprop(g=gf_net,input_m = input_m, no_cores=n_cores, r=0.1)
#Propagated for the network fragment-fragment
ff_prop=rwr_OVprop(g=ff_net,input_m = gf_prop, no_cores=n_cores, r=0.8)

```

# Network visualization
Using the obtained propagation scores, we can create the final network and visualize it using a graphic interface.
```{r network visualization}
#Create igraph object with all the information included
net=create_net2plot(gf_net,input_m,gf_prop,ann_net_b,frag_pattern="frag",ff_net,ff_prop)

#Start GUI
start_GUI(net, ann_net_b, chr_len)
```
 









